---
title: "Data Wrangling with Tidyverse"
author: "Manisha Barse"
date: "2025-05-30"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("nycflights13")  # for data
library("tidyverse")     # for data manipulation (includes dplyr, tidyr)
library("sessioninfo")
```

## Check the structure of the flights dataset
```{r}
glimpse(flights)
```


# 1. RESHAPING DATA


## Delays by month and carrier
```{r}
monthly_delays <- flights |>
  filter(!is.na(arr_delay)) |>
  group_by(month, carrier) |>
  summarise(
    avg_delay = mean(arr_delay),
    .groups = "drop"
  )

print(monthly_delays)
# This is a "long" format: each row = a month-carrier pair
```

## pivot_wider use case
### A table that has one row per month and one column per carrier's average delay
```{r}
wide_delays <- monthly_delays |>
  pivot_wider(
    names_from = carrier,
    values_from = avg_delay
  )

print(wide_delays)
```

## Reverse operation: wide table back to long
```{r}
long_delays <- wide_delays |>
  pivot_longer(
    cols = -month,
    names_to = "carrier",
    values_to = "avg_delay"
  )

print(long_delays)
# collapses multiple columns into key-value pairs
```

## separate() example
```{r}
datetime_example <- flights |>
  select(time_hour) |>
  slice(1:5) |>
  mutate(time_hour_char = as.character(time_hour))

print(datetime_example)

# separate() splits a column into multiple columns by a separator
datetime_separated <- datetime_example |>
  separate(
    col = time_hour_char,
    into = c("date", "time"),
    sep = " "
  )

print(datetime_separated)
```

## unite() example
```{r}
flights |> select(year, month, day)

date_united <- flights |>
  select(year, month, day) |>
  slice(1:5) |>
  unite(
    col = "date",
    year, month, day,
    sep = "-"
  )

print(date_united)

# Leading zeros for formatting
date_united <- flights |>
  slice(1:5) |>
  mutate(
    year = as.character(year),
    month = sprintf("%02d", month),
    day = sprintf("%02d", day)
  ) |>
  unite(col = "date", year, month, day, sep = "-") |>
  select(date)

print(date_united)
```

### Summary:
- `pivot_wider()`: Long -> Wide (spread key-value pairs into multiple columns)
- `pivot_longer()`: Wide -> Long (collapse multiple columns into key-value pairs)
- `separate()`: Split one column into multiple columns by separator
- `unite()`: Combine multiple columns into one column, joining by separator

# 2. JOINING TABLES


## airlines data
```{r}
print(head(airlines))
```

## flights data (select a few columns)
```{r}
print(head(flights |> select(year, month, day, carrier, flight)))
```

## INNER JOIN: keep only rows that have matching keys in both tables
```{r}
inner_join_example <- flights |>
  select(year, month, day, carrier, flight) |>
  inner_join(airlines, by = "carrier") |>
  slice(1:10)

print(inner_join_example)
```

## error case (missing 'by' argument)
```{r, eval=FALSE}
flights |>
  select(year, month, day, flight) |>
  inner_join(airlines) # use 'by' and common column

# Error in `inner_join()`:
# ! `by` must be supplied when `x` and `y` have no common
#   variables.
# ℹ Use `cross_join()` to perform a cross-join.
```

## LEFT JOIN: keep all rows from left table, add matching columns from right
```{r}
left_join_example <- flights |>
  select(year, month, day, carrier, flight) |>
  left_join(airlines, by = "carrier") |>
  slice(1:10)

print(left_join_example)
anyNA(left_join_example)

flights_extra <- flights |>
  select(year, month, day, carrier, flight) |>
  slice(1:5) |>
  mutate(carrier = c("AA", "ZZ", "DL", "XX", "UA"))

result <- flights_extra |>
  left_join(airlines, by = "carrier")

print(result) ## NAs present

## replace NAs
result <- flights_extra |>
  left_join(airlines, by = "carrier") |>
  mutate(name = replace_na(name, "Unknown carrier"))

print(result)
```

## RIGHT JOIN: keep all rows from right table, add matching columns from left
```{r}
right_join_example <- flights_extra |>
  select(year, month, day, carrier, flight) |>
  right_join(airlines, by = "carrier")

print(head(right_join_example))

right_join_example <- flights |>
  select(year, month, day, carrier, flight) |>
  right_join(airlines, by = "carrier")

print(head(right_join_example))
```

## FULL JOIN: keep all rows from both tables, fill NA where no match
```{r}
full_join_example <- flights_extra |>
  select(year, month, day, carrier, flight) |>
  full_join(airlines, by = "carrier")

print(head(full_join_example))
```
### Comparing to Base r functions: uses `merge()`
```{r}
df <- as.data.frame(flights)
al <- as.data.frame(airlines)

# Inner join
inner_merge <- merge(df, al, by = "carrier")

# Left join
left_merge <- merge(df, al, by = "carrier", all.x = TRUE)
head(left_merge)
# Right join
right_merge <- merge(df, al, by = "carrier", all.y = TRUE)
head(right_merge)
# Full join
full_merge <- merge(df, al, by = "carrier", all = TRUE)

# Tidyverse joins
inner_join_result <- inner_join(df, al, by = "carrier")
left_join_result <- left_join(df, al, by = "carrier")
right_join_result <- right_join(df, al, by = "carrier")
full_join_result <- full_join(df, al, by = "carrier")
```


## FILTERING JOINS
```{r}
semi_join_example <- flights_extra |>
  select(year, month, day, carrier, flight) |>
  semi_join(airlines, by = "carrier") |>
  slice(1:10)

print(semi_join_example)

anti_join_example <- flights_extra |>
  select(year, month, day, carrier, flight) |>
  anti_join(airlines, by = "carrier") |>
  distinct(carrier)

print(anti_join_example)

anti_join_example <- flights |>
  select(year, month, day, carrier, flight) |>
  anti_join(airlines, by = "carrier") |>
  distinct(carrier)

print(anti_join_example) # empty
```

## Base R equivalent
```{r}
df <- as.data.frame(flights)
al <- as.data.frame(airlines)

# Inner join
inner_merge <- merge(df, al, by = "carrier")

# Left join
left_merge <- merge(df, al, by = "carrier", all.x = TRUE)
head(left_merge)

# Right join
right_merge <- merge(df, al, by = "carrier", all.y = TRUE)
head(right_merge)

# Full join
full_merge <- merge(df, al, by = "carrier", all = TRUE)

# Tidyverse joins
inner_join_result <- inner_join(df, al, by = "carrier")
left_join_result <- left_join(df, al, by = "carrier")
right_join_result <- right_join(df, al, by = "carrier")
full_join_result <- full_join(df, al, by = "carrier")
```

# 3. BINDING

## Bind rows
```{r}
flights_jan <- flights |>
  filter(month == 1) |>
  slice(1:5)

flights_feb <- flights |>
  filter(month == 2) |>
  slice(1:5)

bind_rows_example <- bind_rows(flights_jan, flights_feb)
print(bind_rows_example)

flights_jan_extra <- flights_jan |>
  mutate(extra_col = "extra")  # add a new column

# Bind rows with different columns
bind_rows_extra <- bind_rows(flights_jan, flights_jan_extra) |>
  slice(1:5)
bind_rows_extra$extra_col
```

## Bind columns
```{r}
df1 <- flights_jan |> select(year, month, day)
df2 <- flights_jan |> select(carrier, flight, tailnum)

bind_cols_example <- bind_cols(df1, df2)
print(bind_cols_example)
```

## Comparing to Base r functions
### rbind: works only if both data frames have exactly the same columns
```{r}
rbind_example <- rbind(flights_jan, flights_feb)
print(rbind_example)

# rbind(flights_jan, flights_jan_extra) ## error
# Error in rbind(deparse.level, ...) : 
#   numbers of columns of arguments do not match
bind_rows_example2 <- bind_rows(flights_jan, flights_jan_extra)
print(bind_rows_example2)
anyNA(bind_rows_example2)
```

### cbind, use carefully
```{r}
cbind_example <- cbind(df1, df2)
print(cbind_example)

df_short <- flights_jan |> select(carrier) |> slice(1:3)
# cbind_mismatch <- cbind(df1, df_short) ## error
# Error in data.frame(..., check.names = FALSE) : 
#   arguments imply differing number of rows: 5, 3
# bind_cols(df1, df_short) #error

cbind(a = 1:5, b = 1:3)# recycles: caution
# bind_cols(a=1:5,b=1:3) # error
# Error in `bind_cols()`:
# ! Can't recycle `a` (size 5) to match `b` (size 3).
```

# 4. SET OPERATORS

```{r}
flights1 <- flights |>
  filter(month == 1, day <= 10) |>
  select(year, month, day, carrier, flight)

flights2 <- flights |>
  filter(month == 1, day >= 5, day <= 15) |>
  select(year, month, day, carrier, flight)

# INTERSECT: rows common to both
intersect_example <- intersect(flights1, flights2)
print(intersect_example)

# UNION: all unique rows from both
union_example <- union(flights1, flights2)
print(union_example)

# SET DIFF: rows in flights1 not in flights2
setdiff_example <- setdiff(flights1, flights2)
print(setdiff_example)

# SET EQUAL: check if sets are identical
setequal_example <- setequal(flights1, flights2)
print(setequal_example)
```

## Reproducibility information
```{r}
print("Reproducibility information:")
Sys.time()
proc.time()
options(width = 120)
session_info()
```
